---
title: "Vacancy rates and rent change (xkcd style)"
author: "Jens von Bergmann"
date: "18/02/2022"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
	dpi=300
)
```

This graph is based on [CMHC Rms data](https://doodles.mountainmath.ca/blog/2022/02/18/vacancy-rates-and-rent-change-2021-update/), combining Edmonton and Calgary to reduce noise. Edmonton and Calgary are good examples for highlighting the relationship between rents and vacancy rates because they don't have rent control, fairly elastic housing supply, and have gone through several economic boom-bust cycles.

The rent change is fixed-sample rent change adjusted for inflation. Fixed sample means that for each cycle it compares rents across the same set of physical units, avoiding it getting skewed by compositional effects like a lot of new units coming online in a particular year.

The code for the graph is [available for anyone to reproduce or adapt](https://github.com/mountainMath/xkcd_exponential/blob/main/rental_xkcd.Rmd).


```{r fig.height=6, fig.width=6}
library(tidyverse)
library(xkcd)
library(cmhc)
library(cansim)

set.seed(123)

# generate the exponential curves with random noise
get_timeseries <- function(table_id,region_params) {
  d1 <- get_cmhc(cmhc_timeseries_params(table_id = table_id,
                                        region=region_params,
                                        filter = list(season="October"))) %>%
    mutate(Month="October")
  d2 <- get_cmhc(cmhc_timeseries_params(table_id = table_id,
                                        region=region_params,
                                        filter = list(season="April"))) %>%
    mutate(Month="April")
  bind_rows(d1,d2) %>% 
    rename(Date=X1) %>%
    mutate(Date=str_extract(Date,"\\d{4}")) %>%
    mutate(Date=as.Date(paste0(Date," ",Month," 01"),format="%Y %B %d")) %>%
    select(-Month)
}

region_params <- cmhc_region_params_from_census("48825")
region_params2 <- cmhc_region_params_from_census("48835")
bedroom_vars <- c("Bachelor", "1 Bedroom", "2 Bedroom", "3 Bedroom +", "Total")

inflation <- get_cansim_vector("v108785713","1990-01-01") %>%
  rename(CPI=val_norm) %>%
  select(Date,CPI) %>%
  mutate(CPI=CPI/100)

d1 <- bind_rows(
  dat_vacancy <- get_timeseries(cmhc_table_list$`Rms Vacancy Rate Time Series`,region_params) %>%
    mutate(Series="Vacancy rate"),
  get_timeseries(cmhc_table_list$`Rms Rent Change Time Series`,region_params) %>%
    mutate(Series="Rent change")) %>%
  mutate(name="Calgary")

d2 <- bind_rows(
  dat_vacancy <- get_timeseries(cmhc_table_list$`Rms Vacancy Rate Time Series`,region_params2) %>%
    mutate(Series="Vacancy rate"),
  get_timeseries(cmhc_table_list$`Rms Rent Change Time Series`,region_params2) %>%
    mutate(Series="Rent change")) %>%
  mutate(name="Edmonton")

data <-   bind_rows(d1,d2) %>%
  standardize_cmhc_data() %>%
  mutate(CV=case_when(Quality=="a" ~ 1,
                      Quality=="b" ~ 3,
                      Quality=="c" ~ 6,
                      Quality=="d" ~ 8,
                      TRUE ~ 15)) %>%
  mutate(Value=Value/100) %>%
  group_by(Series,Date,Bedrooms) %>%
  summarize(Value=weighted.mean(Value,w=1/CV,na.rm=TRUE),
            CV=weighted.mean(CV,w=1/CV,na.rm=TRUE),
            .groups="drop") %>%
  group_by(Series,Date) %>%
  mutate(mean=weighted.mean(Value,w=1/CV,na.rm=TRUE)) %>%
  left_join(inflation,by="Date") %>%
  mutate(name="Calgary/Edmonton")

reference_date <- as.Date("2000-01-01")

plot_data <- data %>% 
  filter(Bedrooms=="Total") %>%
  mutate(Value=ifelse(Series=="Rent change",Value-CPI,Value)) %>%
  group_by(Series,name,Bedrooms) %>%
  mutate(Next=lead(Value,order_by = Date)) %>%
  mutate(Previous=lag(Value,order_by = Date)) %>%
  mutate(diff=difftime(Date,lag(Date,order_by = Date)) %>% as.integer) %>%
  ungroup %>%
  mutate(keep=!is.na(Value)| !(diff<200 & !is.na(Previous))) %>%
  filter(keep) %>%
  mutate(Value=coalesce(Value,mean)) %>%
  mutate(day=difftime(Date,reference_date) %>% as.integer) %>%
  select(Date,day,Series,Value) 


l1 <- "Everyone else"
l2 <- "Critical urban geographers"

text_items <- tibble(Type=factor(c(l1,l1,l2,l2),
                                 levels=c(l1,l2)),
                     Date=as.Date(c("2015-01-01","2001-01-01","2015-01-01","2001-01-01")),
                     Value=c(0.18,-0.06,0.18,-0.06),
                     label=c("Rents rise when\nvacancy rates are low",
                             "Rents fall when\nvacancy rates are high",
                             "Financialization\n& gentrification",
                             "altruistic\nmom & pop landlords")) %>%
                     # label=c("Rents rise when\nvacancy rates are low",
                     #         "Rents fall when\nvacancy rates are high",
                     #         "Landlords are greedy\nand commodified",
                     #         "Landlords are altruistic\nand decommodified")) %>%
  mutate(day=difftime(Date,reference_date) %>% as.integer)


             
curve_data_calgary <- tibble(x=as.Date(c("2011-01-01","2011-01-01","2015-01-01","1996-10-01","2005-05-01","2005-05-01")),
                     xend=as.Date(c("1998-12-01","2006-12-01","2013-10-01","1993-10-01","2010-02-01","2016-08-01")),
                     Value=c(0.18,0.18,0.15,-0.07,-0.07,-0.07),
                     Valueend=c(0.135,0.16,0.075,-0.04,-0.074,-0.09),
                     curvature = c(0.1,0.1,-0.1,-0.2,0.1,0.1)) %>%
  mutate(day=difftime(x,reference_date) %>% as.integer,
         dayend=difftime(xend,reference_date) %>% as.integer)
curve_data_ce <- tibble(x=as.Date(c("2011-01-01","2011-01-01","2015-01-01","1996-06-01","2005-09-01","2005-09-01")),
                     xend=as.Date(c("1998-12-01","2007-12-01","2014-10-01","1994-10-01","2010-01-01","2016-07-01")),
                     Value=c(0.18,0.18,0.15,-0.07,-0.07,-0.07),
                     Valueend=c(0.085,0.16,0.07,-0.04,-0.065,-0.075),
                     curvature = c(0.1,0.1,-0.1,-0.2,0.1,0.1)) %>%
  mutate(day=difftime(x,reference_date) %>% as.integer,
         dayend=difftime(xend,reference_date) %>% as.integer)

baselines <- tibble(Series=c("Rent change","Vacancy rate"),
                    Value=c(0,0.03))

mapping <- aes(x,  y,
                scale,ratioxy,angleofspine,
               anglerighthumerus,anglelefthumerus,anglerightradius,
               angleleftradius,angleleftleg,anglerightleg,angleofneck)

dataman <- data.frame( x= as.integer(difftime(as.Date("1992-01-01"),reference_date)), 
                       y=0.2,
                       scale = 0.05 ,
                       ratioxy = 15000,
                       angleofspine =  -pi/2  ,
                       anglerighthumerus = c(-pi/6, -pi/6),
                       anglelefthumerus = c(-pi/2 - pi/6, -pi/2 - pi/6),
                       anglerightradius = c(pi/5, -pi/5),
                       angleleftradius = c(pi/5, -pi/5),
                       angleleftleg = 3*pi/2  + pi / 12 ,
                       anglerightleg = 3*pi/2  - pi / 12,
                       angleofneck = runif(1, 3*pi/2-pi/10, 3*pi/2+pi/10),
                       Type=c(l2,l1))


g <- ggplot(plot_data,aes(x=day,y=Value))+
  geom_hline(data=baselines,aes(linetype=Series,yintercept=Value),size=0.5) +
  geom_line(aes(linetype=Series,size=Series)) +
  theme_xkcd() +
  scale_size_manual(values=c("Rent change"=1,"Vacancy rate"=0.75),
                    guide="none") +
  theme(legend.position = "bottom") +
  geom_point(size=1.2,shape=21) +
  scale_y_continuous(labels=scales::percent,breaks=seq(-0.05,0.15,0.05)) +
  scale_x_continuous(breaks=NULL,minor_breaks = NULL) +
  labs(x=NULL,y=NULL,
       #caption="MountainMath, Data: CMHC Rms (Calgary)",
       linetype=NULL) +
  geom_text(data=text_items,family="xkcd",
            aes(label=label)) +
  scale_linetype_manual(values=c("Rent change"="solid","Vacancy rate"="dashed")) +
  theme(panel.spacing.y = unit(2, "lines")) +
  xkcdman(mapping,dataman,inherit.aes=FALSE) +
  facet_wrap(~Type, ncol=1) +
  geom_curve(data=curve_data_ce[c(1),],
             aes(x=day,xend=dayend,yend=Valueend),
             arrow = arrow(length=unit(0.25,"cm")),
             curvature = 0.2) +
  geom_curve(data=curve_data_ce[c(2,5,6),],
             aes(x=day,xend=dayend,yend=Valueend),
             arrow = arrow(length=unit(0.25,"cm")),
             curvature = 0.1) +
  geom_curve(data=curve_data_ce[3,],
             aes(x=day,xend=dayend,yend=Valueend),
             arrow = arrow(length=unit(0.25,"cm")),
             curvature = -0.1) +
  geom_curve(data=curve_data_ce[4,],
             aes(x=day,xend=dayend,yend=Valueend),
             arrow = arrow(length=unit(0.25,"cm")),
             curvature = -0.2) 

g
```




